\texttt{ensmallen}: a flexible C++ library for efficient function optimization

### 2018-10-15 doc/mlsys_paper/paper.tex
notes on why this section should be removed

\vspace*{-0.3em}
\section{Available Optimizers}
\vspace*{-0.5em}

Due to the straightforward abstraction framework, {\tt ensmallen} is able to
provide a large set of diverse optimization algorithms.  Below is a list of
currently available optimizers:

% I guess to save some space we should group them.
%% CS: should we add citations to some (or all) of the listed optimisers?
%% CS: the MLSYS CFP states that "papers can be up to 6 pages (not including references)",
%% CS: so technically we can have 10000+ references.
\vspace*{-0.4em}
\begin{itemize}
  \item {\bf SGD variants:} Stochastic Gradient Descent (SGD), Stochastic
      Coordinate Descent (SCD), Parallel Stochastic Gradient Descent (Hogwild!),
      Stochastic Gradient Descent with Restarts (SGDR), SMORMS3, AdaGrad,
      AdaDelta, RMSProp, Adam, AdaMax

@@

%% CS: "in development" is ambiguous here;
%% CS: is it in development by AI/ML researchers in general,
%% CS: or by ensmallen developers/contributors?
% In addition, many optimization algorithms are currently in development and will
% be released in the future.

%% CS: less ambiguous proclamation that has similar effect:
We aim to expand {\tt ensmallen} with further optimization techniques
as the need arises.  Since {\tt ensmallen} is open source,
external contributions to the codebase are welcome.

\vspace*{-0.3em}
\section{Templates for speed}
\vspace*{-0.5em}

Many optimization toolkits depend on the user passing in a function pointer to a
function that needs to be optimized.  For instance, the {\tt scipy.optimize}
package in Python can be used like this:

@@

During the call to {\tt minimize()}, the {\tt rosen()} function will be called
repeatedly---but it will be called as a function pointer, which must be
dereferenced each time {\tt rosen()} is called.  In addition, some
optimizations such as inlining (and many optimizations enabled by inlining) are
not possible when using function pointers.  Since some optimization algorithms
need to compute the objective very many times, the overhead of the lookup and
missed optimization opportunities may be non-negligible.  This applies even for
SGD-like optimizers that train machine learning models on very large datasets!

Since {\tt ensmallen} is written in C++ with templates, the function pointer
dereferencing overhead can be avoided and the compiler can choose to inline the
function call and perform additional optimizations when possible, such as
temporary variable elimination and dead code pruning.

>>>>

\vspace*{-0.3em}
\section{Available optimizers}
\vspace*{-0.5em}

Due to the easy abstraction, {\tt ensmallen} is able to provide a large set of
diverse optimization algorithms.  Below is a list of what is currently
available.

% I guess to save some space we should group them.
\vspace*{-0.4em}
\begin{itemize}
  \item {\bf SGD variants:} Stochastic Gradient Descent (SGD), Stochastic
      Coordinate Descent (SCD), Parallel Stochastic Gradient Descent (Hogwild!),
      Stochastic Gradient Descent with Restarts (SGDR), SMORMS3, AdaGrad,
      AdaDelta, RMSProp, Adam, AdaMax

@@

In addition, many optimization algorithms are currently in development and will
be released in the future.

\vspace*{-0.3em}
\section{Templates for Speed}
%\section{Automatic Metaprogramming for Efficiency Gains}
\vspace*{-0.5em}

%% CS: this sections is very weak and not convincing; it basically states
%% CS: "look, we can do inlines", but provides no real evidence
%% CS: and/or citations where the supposed efficiency gains are demonstrated.
%% CS: i suggest removing this section entirely,
%% CS: as it causes more harm than good (from a reviewer's point of view).
%% 
%% CS: in practice, effect of C++ function inlining is only notably
%% CS: pronounced when dealing with data comprised of lots of small vectors.
%% 
%% CS: for this section to be fleshed out, it would require reasoning based
%% CS: on citations covering compiler optimisation, effect of data size,
%% CS: experiments showing the effect of various vector sizes
%% CS: (eg. with and without inlining).
%% CS: there is no room in this paper to do this properly,
%% CS: and doing a poor job will reduce likelihood of acceptance.

Many optimization toolkits depend on the user providing a pointer to the
function that needs to be optimized.  For instance, the {\tt scipy.optimize}
package in Python can be used like this:

@@

During the call to {\tt minimize()}, the {\tt rosen()} function will be called
repeatedly---but it will be called as a function pointer, which must be
dereferenced each time {\tt rosen()} is called.  In addition, some
optimizations such as inlining (and many optimizations enabled by inlining) are
not possible when using function pointers.  Since many optimization algorithms
need to compute the objective numerous times, the overhead of the lookup and
missed optimization opportunities may be non-negligible.  This applies even for
SGD-like optimizers that train machine learning models on very large datasets.

%% CS: this is uninteresting and basically looks like filler
Since {\tt ensmallen} is written in C++ with templates, the function pointer
dereferencing overhead can be avoided and the compiler can choose to inline the
function call and perform additional optimizations when possible, such as
temporary variable elimination and dead code pruning.
